---
title: "AQ project"
author: "Claudia Lopez & Vanessa Ortiz"
date: "2024-04-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#intro



#Data 

What’s interesting about this data is that we can evaluate the relationship between air pollution, and death rates. Not only that but we can find out the largest factor of air pollution and why the air is of such bad quality in the Middle East. The reason for choosing this data and topic is because it sparks an interest in the health of our planet, and ultimately can help find a solution to the problem.

```{r}
library(tidyverse)

#both data sets contain more than 50k entries with many updates throughout a singular year
data <- read_csv2("openaq.csv")
WAQ_data <- read.csv2("world_air_quality.csv") 

#simple data yet specifies the different levels of air quality in each country 
GAP <- read.csv("global air pollution dataset.csv")

DR_data <- read.csv("death_rate_airpollution.csv")

#this data set has the amount and types of pollution each country has contributed from 2010-2020          
types_pollution <- read.csv("country_level_data_0.csv")
#specifies what causes the emission and what is the main pollutant in the air from 2000-2020 in each country
T_Emmission <- read.csv("Total Emissions Per Country (2000-2020).csv")
```



#Questions 

1. what are the factors of air pollution?
 - how do these factors vary through out each continent? 
2. Were there any changes in factors or levels of pollution before and after Covid-19?
3. Is there a noticeable difference in walkable citites in contrast to non-walkable cities?
4. What is the death rate caused by air pollution in each continent/country?
5. Why does the Middle East have the worst air quality 